---
title: "Cohorts, Samples and Sample Sets"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cohorts, Samples and Sample Sets}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
resource_files:
  - '../man/figures/cohorts.svg'
  - '../man/figures/samples.svg'
  - '../man/figures/sample_sets.svg'
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(quincunx)
```

## Cohorts

A cohort is a group of individuals with a shared characteristic. Cohorts are
identified in quincunx by the `cohort_symbol` variable. See
`vignette('getting-cohorts')` on how to find associated polygenic scores.

```{r  out.width = "90%", echo=FALSE}
knitr::include_graphics('../man/figures/cohorts.svg') 
```

Using `get_cohorts()` to retrieve associated PGS identifiers with cohort
`"PROMIS"`:

```{r}
get_cohorts('PROMIS')
```

## Samples

A sample is a group of participants associated with none, one or more catalogued
cohorts. The selection from a cohort can be either a subset or its totality.
Samples are not identified in PGS Catalog with a global unique identifier, but
quincunx assigns a surrogate identifier (`sample_id`) to allow relations between
tables.

```{r  out.width = "90%", echo=FALSE}
knitr::include_graphics('../man/figures/samples.svg') 
```

Sample composition is provided in slot `cohorts` from objects `scores` returned
by the `get_scores()` function.

```{r}
library(dplyr, warn.conflicts = FALSE)

# PGS000001 is one of the polygenic scores that is based upon participants from
# cohort PROMIS
pgs_11 <- get_scores('PGS000011')

# Cohort PROMIS is included in sample no. 2, along with LOLIPOP
filter(pgs_11@cohorts, sample_id == 2L)
```

To know a few more details about samples, look into the `samples` slot of the
object `scores`:

```{r}
filter(pgs_11@samples, sample_id == 2L)
```

## Sample sets

A sample set is a group of samples used in a polygenic score evaluation. Each
sample set is identified in the PGS Catalog by a unique sample set identifier
(`pss_id`).

```{r  out.width = "90%", echo=FALSE}
knitr::include_graphics('../man/figures/sample_sets.svg') 
```

To find the sample sets that included a specific cohort, we start by getting
the PGS identifiers associated with a cohort, e.g. MHI:

```{r}
# Note that by the definition of sample set, samples included in sample sets
# are only used at PGS evaluation stages.
filter(get_cohorts('MHI')@pgs_ids, stage == 'eval')
```

PGS000013 is one of the polygenic scores whose evaluation used participants from the cohort
MHI. We retrieve now the sample sets used in the evaluation of PGS000013:

```{r}
# Sample sets used in the evaluation of the PGS000013
pgs_13_sset <- get_sample_sets(pgs_id = 'PGS000013')
glimpse(pgs_13_sset@sample_sets)
```

One of the sample sets used to evaluate PGS000013 is PSS000020. We can retrieve
a `sample_set` object that contains its composition, i.e., the samples and
cohorts included, along with other details:

```{r}
get_sample_sets('PSS000020')
```
