---
title: "Getting PGS Publications"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting PGS Publications}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## PGS Publications

PGS publications are those published works that underlie the data provided by
the PGS Catalog.

To retrieve these publications use the function `get_publications()`. You may
use one of the following search criteria (or a combination thereof):

- `pgp_id`: the publication identifier assigned by the PGS Catalog;
- `pgs_id`: the polygenic score identifier;
- `pubmed_id`: PubMed identifier;
- `author`: an author last name.

If you do not pass any of the criteria above, then `get_publications()` will
retrieve all publications in the Catalog.

## Getting PGS Publications

Getting PGS publications by their identifiers:

```{r}
library(quincunx)

get_publications(pgp_id = c('PGP000001', 'PGP000002'))
```

By `pgs_id`:

```{r}
library(quincunx)

get_publications(pgs_id = 'PGS000003')
```

By `pubmed_id`:

```{r}
library(quincunx)

get_publications(pubmed_id = '30554720')
```

By `author`:

```{r}
library(quincunx)

get_publications(author = 'Natarajan')
```

## Getting PGS Publications by other criteria

The PGS Catalog REST API only supports searches by those criteria mentioned above. If you would like to get results by other criteria, e.g., `publication_date`, then you need to retrieve all publications and filter them afterwards.

As an example, let's download all publications and then keep only those publications published in 2021:

```{r}
all_pub <- get_publications(interactive = FALSE)
all_pub@publications
```

Filtering based on the year 2021:

```{r}
library(dplyr, warn.conflicts = FALSE)

# Determine the PGP ids whose publication date falls within 2021.
pgp_ids_2021 <-
  filter(
    all_pub@publications,
    publication_date >= '2021-01-01' &
      publication_date <= '2021-12-31'
  ) %>%
  pull('pgp_id')

# Filtering based on the PGP ids
pub_2021 <- all_pub[pgp_ids_2021]

# Print the first 10 PGS publications
pub_2021@publications
```
