---
title: "Getting PGS Performance Metrics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting PGS Performance Metrics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## PGS performance metrics

Performance metrics assess the validity of a PGS in a Sample Set. This
assessment is performed on samples not used for score development.

Performance metrics are retrieved with the function `get_performance_metrics()`.
The returned data is provided as an S4 object of class `performance_metrics`.

Common metrics include:

- standardized effect sizes: odds ratios or hazard ratios, and regression
coefficients $\beta$, see slot `pgs_effect_sizes`;
- classification accuracy metrics: area under the receiver operating
characteristic curve, C-index and area under the precision-recall curve, see
slot `pgs_classification_metrics`;
- other relevant metrics: calibration ($\chi^2$)), see slot `pgs_other_metrics`.

The covariates used in the model (most commonly age, sex and genetic principal
components to account for the population structure) are also recorded for each
set of metrics. These can be found in the slot `demographics`.

## Getting PGS performance metrics

In the PGS Catalog, performance metrics have been catalogued and have an
associated identifier that starts with the prefix `"PPM"`. To retrieve the
performance metrics associated with one assessment of a polygenic score, you can
use directly its identifier:

```{r}
library(quincunx)

get_performance_metrics(ppm_id = 'PPM000001')
```

## Searching by PGS identifier

Alternatively, you could also search by the associated PGS identifier, i.e.
`"PGS000001"`:

```{r}
get_performance_metrics(pgs_id = 'PGS000001')
```

As you can see, when you search by `'PGS000001'`, we get multiple PPM
identifiers (PPM000001 included). This is because a PGS could have been assessed
multiple independent times, each assessment resulting in its own performance
metrics data entry, with its own associated identifier.

## Vectorised search

The function `get_performance_metrics()` is vectorised over `ppm_id` and
`pgs_id` and you could readily retrieve performance metrics for a set of
polygenic scores by providing a vector of identifiers (e.g. PGSes 42 thru 46):

```{r}
ppm <- get_performance_metrics(pgs_id = sprintf("PGS%06d", 42:46))
print(ppm@performance_metrics, n = Inf)
```
