---
title: "Getting Traits"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Traits}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
notes-after-punctuation: false
bibliography: "`r system.file('references.bib', package = 'quincunx')`"
csl: "`r system.file('nature.csl', package = 'quincunx')`"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(quincunx)
```

## Introduction

Polygenic scores (PGSs) are annotated with information about the phenotype that
it predicts, i.e. the reported trait (as reported in the original publication).
This can be found as the column `reported_trait` in slot `scores` of `scores`
objects:

```{r}
pgs_01 <- get_scores('PGS000001')
pgs_01@scores
```

The predicted phenotype is also mapped to Experimental Factor Ontology (EFO)
terms (a controlled vocabulary for the unambiguous identification of traits and
diseases, and their relationships), namely, the EFO trait. The EFO traits
associated with a polygenic score can also be found in `scores` objects in the
slot `traits`, column `trait`:

```{r}
pgs_01@traits
```

Many PGSs have been developed and demonstrated to be predictive of common
complex traits, e.g. body mass index (BMI) [@khera2019], blood lipids
[@kuchenbaecker2019] and educational attainment [@lee2018].

Similarly, PGSs for various diseases have been shown to be predictive of disease
incidence, defining marked increases in risk over the life course or at earlier
ages for people with high PGSs, e.g. coronary artery disease
[@inouye2018;@khera2018], breast cancer [@mavaddat2019] and schizophrenia
[@zheutlin2019].

## Getting catalogued traits from PGS Catalog

If you are interested in retrieving polygenic scores from the Catalog, you might
want to search them by the trait they predict. `get_scores()` is the function
that searches for `PGSs`, however, this function only allows to search by
`pgs_id`, `efo_id` or `pubmed_id`. So in order to search by a trait term, we
need to first find the associated EFO identifiers (`efo_id`).

To search for traits (or diseases), you use the function `get_traits()`. With
this function you can search by:

- The EFO trait identifier: `efo_id`;
- or by the trait term: a term to be matched in the EFO identifier (`efo_id`),
label, description synonyms, trait categories, or external mapped terms.

The most useful search criteria is the trait term, and that is typically want
you will want to use. Unless you already know the EFO trait you are interested
in, and are looking for extra details about it, you won't search directly with
the EFO identifier.

### Basic example

Let's say you are interested in PGSs related to medical condition, stroke. Then
you can search for `"stroke"` with `get_traits()`:

```{r}
get_traits(trait_term = 'stroke')
```

As can be seen from the returned `traits` object, we get a set of six tables
(slots) that include several details about stroke.

In the first table `traits` we got only one row, indicating that this query
returned only one trait in the Catalog. This trait is named `"stroke"` (column
`trait`), and is unambiguously identified by the EFO identifier EFO_0000712.

### Exact matching

By default, the trait term is matched exactly. If you want to relax the
matching, then indicate with the parameter `exact_term` set to `FALSE`. This way
you will get, potentially, more results, in this example case, ischemic stroke
(HP_0002140) is now also returned:

```{r}
get_traits(trait_term = 'stroke', exact_term = FALSE)
```

### Subtraits (child traits)

By default, subtraits (child traits), are not retrieved by `get_traits()`. If
you want to get all matching traits and those that are child traits thereof,
then indicate with the parameter `include_children` set to `TRUE`. Here is an
example with `"breast cancer"`:

```{r}
get_traits(trait_term = 'breast cancer', include_children = TRUE)
```

The column `is_child` indicates whether that trait is being retrieved because it
is a direct result of the query or not. `is_child` is `TRUE` when the trait is
returned because it is a child trait of a matching trait, and `FALSE` if a
direct result of the query.

In the case of child traits, the column `parent_efo_id` indicates the EFO trait
identifier of the parent trait, i.e. the direct matching trait, or `NA`
otherwise.

## Getting all traits

To retrieve all traits simply leave the parameters `efo_id` and `trait_term` as
`NULL` (default):

```{r eval=FALSE}
get_traits()
```

`r if (knitr::is_html_output()) '## References'`
