---
title: "Getting Cohorts"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Cohorts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Cohorts

A cohort is a group of individuals with a shared characteristic. Cohorts are
identified in quincunx by the `cohort_symbol` variable. Participants in cohorts
are used to define samples, which in turn, are used to assemble sample sets.
For more details on the relationship between the concepts of cohorts, samples
and sample sets, see `vignette('cohorts-samples-sample-sets')`.

Given that study participants typically come from one or more catalogued cohorts
and that cohorts can have a strong bias ancestry composition --- i.e., most
cohorts are mostly composed of European-ancestry individuals ---, it can be
really important to know which cohorts have been used at the different stages of
a Polygenic Score (PGS) life cycle to assess the transferability of PGS
performance.

## Getting cohorts

If you know before hand the cohort acronyms (e.g., `"23andMe"`) that you are
interested in, then you can get their full name and associated PGS identifiers
using the `get_cohorts()` function by providing their symbols with the
parameter `cohort_symbol`:

```{r}
library(quincunx)
get_cohorts(cohort_symbol = '23andMe')
```

The `pg_ids` lists associated PGS identifiers with the queried cohorts. The
`stage` variable indicates the PGS stage in which the cohort was used.

To get all catalogued cohorts in the PGS Catalog, leave the `cohort_symbol`
parameter as `NULL` (default). Note that, in this case, it may take a few
minutes for the download to complete.


