---
title: "Contributing to quincunx development"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Contributing to quincunx development}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
resource_files:
  - '../man/figures/quincunx_core_fns_nofonts.svg'
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center"
)
```

If you feel like contributing to quincunx source code please feel free to make a pull request or ask questions in GitHub Issues.

## API underlying retrieval functions

To gain a glimpse of how the core functions of quincunx are organised and dependent on each other please check out this table:
```{r  out.width = "100%", echo=FALSE}
knitr::include_graphics('../man/figures/quincunx_core_fns_nofonts.svg') 
```

### Family of `get_<entities>()` functions

At the exposed API, the family of functions that the user sees are those `get_<entities>()`, e.g., `get_scores()` or `get_sample_sets()`. The family of internal functions `coerce_to_S4_<entities>()` translate a list of relational tibbles to an S4 object. The advantage of having these lists of tibbles encapsulated as S4 objects is that you can then benefit from methods written for them. The placeholder `<entities>` stands for one of: `scores`, `publications`, `sample_sets`, `performance_metrics`, `traits`, `trait_categories`, `cohorts` or `releases`.

### Family of `get_<endpoint>` functions

`<endpoint>` is typically a loosely based snake case name of a PGS Catalog endpoint, e.g., the endpoint `'/rest/score/search'` used with the `trait_id` parameter has the corresponding quincunx function: `get_score_by_trait_id()`.

### Family of `get_<type>` functions

`<type>` is often the name of a PGS Catalog entity in the singular, e.g., `score` for polygenic scores, that corresponds to the function `get_score()`. This family of functions will return always the same type of return value, that is a list of relational tibbles. The heavy lifting of translation from JSON to tibbles happens in `as_tidy_tables_<entities>`.

### Low-level functions: `get()`, `request_all()` and `request()`

At the bottom of the stack `httr::GET()` is the actual function that retrieves a JSON response from a specific URL. `request()` is quincunx's fundamental function after `httr::GET()`. `request()` calls a [memoised](https://github.com/r-lib/memoise) version of `httr::GET()` ---`memoised_GET()`--- to make the package more agile and to avoid hammering the REST API server unnecessarily. `request_all()` does essentially the same as `request()` but handles pagination, meaning that it will follow all pages until it has gathered all results. The `get()` function adds metadata columns to the tibbles, namely, `..resource` (full endpoint URL), `..timestamp` (endpoint retrieval timestamp), `..page` (endpoint pagination).

## S4 classes

S4 classes are defined and documented in source files named `class-<entities>.R`, e.g., `class-scores.R` or `class-performance_metrics.R`

## Parsing of JSON to tibbles

The conversion from JSON use done using the terrific package [tidyjson](https://github.com/colearendt/tidyjson). The source files containing the code for this are `parse-<entities>.R`, e.g., `parse-cohorts.R` or `parse-publications.R`. Some parsing functions are also used across different entities, and in those cases they have dedicated source files: `unwrap_<entities>.R`, e.g., `unwrap_efotrait.R` or `unwrap_sample.R`.
