---
title: "PGS Catalog Releases"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{PGS Catalog Releases}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  fig.align = "center"
)
library(quincunx)
```

The function `get_releases()` gets you information about PGS Catalog releases; namely, the number of newly released polygenic scores, publications and performance metrics, as well as their respective associated identifiers (i.e., new `pgs_ids`, `pgp_ids` and `ppm_ids`).

## Latest release of the PGS Catalog

To get information about the latest release, run `get_releases()` with default argument values:

```{r releases-latest}
get_releases() # Or get_releases(date = 'latest')
```

## Release by date

You can use the argument `date` to select a specific date (if a release exists for this day) in the format `"YYYY-MM-DD"`, e.g., `"2020-11-20"`:

```{r releases-by-date}
get_releases(date = '2020-11-20')
```

## All releases

```{r releases-all}
(all_releases <- get_releases(date = 'all'))
```


```{r releases-over-time, echo=FALSE,message=FALSE}
library(dplyr)
library(ggplot2)
library(plotly)

df <- dplyr::arrange(all_releases@releases, date)
p <- ggplot(df, aes(x = date, y = cumsum(n_pgs))) +
  geom_point() +
  geom_line() +
  scale_x_date(date_labels = '%b %Y', date_breaks = "1 month") +
  scale_y_continuous(breaks = seq(0, 700, 100), limits = c(0, 700)) +
  labs(title = 'PGS Catalog Releases') +
  ylab('Cumulative number of released PGS') +
  theme(axis.text.x = element_text(angle=45, hjust=1),
        axis.title.x=element_blank(),
        axis.ticks.x=element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.border = element_blank()) 
ggplotly(p)
```
