---
title: "PGS Catalog Releases"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{PGS Catalog Releases}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  fig.align = "center"
)
options(tibble.print_min = 4, tibble.print_max = 4)
library(quincunx)
```

The function `get_releases()` gets you information about PGS Catalog releases; namely, the number of newly released polygenic scores, publications and performance metrics, as well as their respective associated identifiers (i.e., new `pgs_ids`, `pgp_ids` and `ppm_ids`).

## Latest release of the PGS Catalog

To get information about the latest release, run `get_releases()` with default argument values:

```{r releases-latest}
get_releases() # Or get_releases(date = 'latest')
```

## Release by date

You can use the argument `date` to select a specific date (if a release exists for this day) in the format `"YYYY-MM-DD"`, e.g., `"2020-11-20"`:

```{r releases-by-date}
get_releases(date = '2020-11-20')
```

## All releases

```{r releases-all}
(all_releases <- get_releases(date = 'all'))
```

